---
title: "module1"
output: html_document
date: "2024-10-29"
---

```{r}
# Set the file path (if necessary)
file_path <- "/Users/emmahe/Documents/GitHub/Module0-t_team9/module1-group-15/data/biomarker-raw.csv"

# Read the CSV file into a dataframe
biomarker_data <- read.csv(file_path)

# Display the top 6 column names of biomarker_data
head(colnames(biomarker_data), 6)
```

# 1.
```{r}
# 1
# Load necessary libraries
library(ggplot2)
library(dplyr)

# Sample two proteins to visualize their distributions before and after log transformation
# Replace with actual column names in your dataset
sample_proteins <- biomarker_data %>%
  select(`E3.ubiquitin.protein.ligase.CHIP`, `CCAAT.enhancer.binding.protein.beta`)

# Plot distributions before log transformation
ggplot(biomarker_data, aes(x = `E3.ubiquitin.protein.ligase.CHIP`)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  ggtitle("Distribution of E3 ubiquitin-protein ligase CHIP (Raw Values)") +
  xlab("E3 ubiquitin-protein ligase CHIP Levels")

ggplot(biomarker_data, aes(x = `CCAAT.enhancer.binding.protein.beta`)) +
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  ggtitle("Distribution of CCAAT/enhancer-binding protein beta (Raw Values)") +
  xlab("CCAAT/enhancer-binding protein beta Levels")

# Log-transform protein levels (add 1 to avoid log(0) if there are zero values)
biomarker_data <- biomarker_data %>%
  mutate(
    `E3.ubiquitin.protein.ligase.CHIP_log` = log(`E3.ubiquitin.protein.ligase.CHIP` + 1),
    `CCAAT.enhancer.binding.protein.beta_log` = log(`CCAAT.enhancer.binding.protein.beta` + 1)
  )

# Plot distributions after log transformation
ggplot(biomarker_data, aes(x = `E3.ubiquitin.protein.ligase.CHIP_log`)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.7) +
  ggtitle("Distribution of E3 ubiquitin-protein ligase CHIP (Log-Transformed)") +
  xlab("Log-Transformed E3 ubiquitin-protein ligase CHIP Levels")

ggplot(biomarker_data, aes(x = `CCAAT.enhancer.binding.protein.beta_log`)) +
  geom_histogram(bins = 30, fill = "green", alpha = 0.7) +
  ggtitle("Distribution of CCAAT/enhancer-binding protein beta (Log-Transformed)") +
  xlab("Log-Transformed CCAAT/enhancer-binding protein beta Levels")
```
Log-transforming the protein levels in biomarker-raw.csv likely aims to address skewed data distributions, where protein values are often right-tailed with extreme outliers. Applying a log transformation makes the distribution more symmetric and reduces the impact of high outliers, helping to stabilize variance and make the data better suited for statistical analysis. This results in a dataset that supports more reliable and accurate modeling, as it aligns better with the assumptions of many analytical methods.

# 2. 






