---
title: "Biomarkers of ASD"
subtitle: "If you want a subtitle put it here"
author: "Cui Christina, Jiajia Feng, Jiahui He, William Mahnke"
date: last-modified
published-title: "Updated"
editor: visual
format: html
code-copy: true
execute:
  message: false
  warning: false
  echo: false
  cache: true
---

Use this as a template. Keep the headers and remove all other text. In all, your report can be quite short. When it is complete, render and then push changes to your team repository.

```{r}
# load any other packages and read data here
library(tidyverse)
```

## Abstract

This report presents a proteomic analysis of serum data aimed at identifying candidate biomarkers for the early detection of autism spectrum disorder (ASD), building on the methodology of Hewitson et al. (2021). We evaluate the effects of various methodological adjustments, including log-transformation of protein levels, outlier handling without data trimming, and the use of a reserved testing partition. To enhance predictive performance, we apply a fuzzy intersection approach for protein selection, constructing refined and alternative panels optimized for classification accuracy. Our findings are compared to previous analyses to assess the robustness and sensitivity of each methodological adjustment.

## Dataset

The dataset includes proteomic measurements from serum samples to identify biomarkers associated with autism spectrum disorder (ASD). It consists of 156 samples and 1,320 variables, each representing different protein levels measured in arbitrary units. Key variables include group classifications (e.g., ASD or TD), specific protein levels, and assessment scores like the ADOS Total Score. Data preprocessing for this analysis involved log-transformation of protein levels, handling outliers without trimming, and splitting into training and testing partitions. This approach enables effective feature selection and classification accuracy assessment for candidate biomarkers.

## Summary of published analysis

The paper combines statistical tests and machine learning to identify key protein biomarkers associated with a specific condition. It starts with univariate T-tests to evaluate each protein’s association with the condition, narrowing down the most significant proteins. Next, Random Forest and Logistic Regression models prioritize these proteins based on feature importance scores and coefficient values, respectively. This dual-model approach helps identify proteins with the greatest predictive potential.

The analysis identified distinct sets of top proteins for each method. For the T-test selection, the key proteins included **PTN, RELT, MAPK2, DERM, Calcineurin, M2-PK, TFF3, FSTL1, CXCL16, MAPK14, Coagulation Factor IX, IgD, MIA, Fas, MMP-2, IGFBP-4, ROR1, Protein S, 14-3-3 protein zeta/delta**, and **TGF-b R III**, yielding an accuracy of **72.05%** on the testing data. The Random Forest model, achieving an accuracy of **78.95%**, selected proteins such as **eIF-4H, MAPK2, PTN, CSK, MAPK14, M2-PK, DERM, Lysozyme, RELT, ILT-4, CD27, Nectin-like protein 2, Coagulation Factor IX, Calcineurin, Notch 1, IgD, IGFBP-1, SOST, GPVI**, and **MMP-2**. For Logistic Regression, the proteins included **SMAD3, HXK1, Myostatin, HIF-1a, CHKB, ISLR2, CSH, HHLA2, RNF43, OAS1, TIMP-1, CD59, SMOC1, 14-3-3 protein beta/alpha, SNP25, LDLR, S100A4, EFNB1, EFNB2**, and **STAT6**. Combining top proteins across methods via a fuzzy intersection improved the classifier accuracy to **82.42%**. This multi-method approach, particularly the fuzzy intersection, achieved the highest overall predictive accuracy.

## Findings

Summarize your findings here. I've included some subheaders in a way that seems natural to me; you can structure this section however you like.

### Impact of preprocessing and outliers

### Question 1

Since `biomarker-raw.csv` contains protein concentration levels across various samples, let's examine the distribution of a sample of these protein values.

```{r}
library(ggplot2)
library(reshape2)
```

```{r}
data <- read.csv("data3/biomarker-raw.csv")

# Convert the selected protein columns to numeric (skip non-numeric values)
proteins <- data[, 3:7]  # Select a sample of 5 protein columns
proteins <- data.frame(lapply(proteins, function(x) as.numeric(as.character(x))))

# Melt the data for ggplot

proteins_long <- melt(proteins, variable.name = "Protein", value.name = "Level")

# Plot histograms for each selected protein
ggplot(proteins_long, aes(x = Level)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_wrap(~Protein, scales = "free") +
  labs(title = "Distribution of Raw Protein Levels (Sample Proteins)",
       x = "Protein Level", y = "Frequency") +
  theme_minimal()

```

The histograms reveal that the distributions of raw protein levels are skewed right, with some values extending to higher ranges. This skewness is a common reason to apply a log transformation, which can help normalize these distributions.\

### Question 2

```{r}
library(dplyr)
library(tidyr)
```

```{r}

library(dplyr)
library(tidyr)


data <- read.csv("data3/biomarker-raw.csv")


protein_data <- data[, -c(1,2)]  


identify_outliers <- function(x) {
  upper_limit <- mean(x, na.rm = TRUE) + 3 * sd(x, na.rm = TRUE)
  lower_limit <- mean(x, na.rm = TRUE) - 3 * sd(x, na.rm = TRUE)
  return(x < lower_limit | x > upper_limit)
}


outliers <- protein_data %>%
  mutate(across(everything(), identify_outliers)) %>%
  rowwise() %>%
  mutate(outlier_count = sum(c_across(everything()), na.rm = TRUE)) %>%
  ungroup()


outlier_summary <- data %>%
  select(Group) %>%
  bind_cols(outliers["outlier_count"])


outlier_by_group <- outlier_summary %>%
  group_by(Group) %>%
  summarise(avg_outliers = mean(outlier_count, na.rm = TRUE),
            total_outliers = sum(outlier_count, na.rm = TRUE),
            subjects_with_outliers = sum(outlier_count > 0))

# View results
print(outlier_by_group)

```

#### Are there specific subjects (not values) that seem to be outliers?

Yes, there are specific subjects with outliers. According to the table of nsummary, there are 2 subjects with outlying values in the ASD group.

#### Are outliers more frequent in one group or the other?

Outliers are more frequent in the ASD group. ASD group with a total of 2 outliers, while the TD group has none.

### Methodlogical variations

### Question 3

```{r}
library(dplyr)
library(caret)  #data partition
library(dplyr)  #ttest
library(purrr)  #ttest
library(broom)  #ttest
library(tidyverse) #ttest
library(rstatix)
library(randomForest) #Random Forest
library(MASS)   #Logistic Regression
```

```{r}
load('data3/biomarker-clean.RData')
head(biomarker_clean)
```

#### Split Training and Testing

```{r}
set.seed(123)

# Split training (80%) and testing (20%) sets
trainIndex <- createDataPartition(biomarker_clean$group, p = 0.8, list = FALSE)
training_data <- biomarker_clean[trainIndex, ]
testing_data <- biomarker_clean[-trainIndex, ]

```

```{r}
head(training_data)
head(testing_data)
```

#### Apply T-test, Random Forest, and Logistic Regression (Using Top 20 Features)

##### T-test Selection

```{r}
library(dplyr)

# Ensure correct names and explicit call to dplyr::select
group_column <- training_data$group  # Confirm column name is correct
protein_columns <- dplyr::select(training_data, -group, -ados)  # Explicitly use dplyr



#T test for all protein
t_test_results <- sapply(protein_columns, function(protein) {
  t.test(protein ~ group_column)$p.value
})

# Convert result into frame
t_test_df <- data.frame(
  protein = names(t_test_results),
  p_value = t_test_results
)

# Extract top 20 proteins
top_proteins_ttest <- t_test_df %>%
  arrange(p_value) %>%
  slice(1:20) %>%
  pull(protein)


print(top_proteins_ttest)
```

##### Random Forest

```{r}

library(randomForest)
library(dplyr)


# Alternative approach if select() conflicts remain
predictors <- training_data[, !names(training_data) %in% c("group", "ados")]
response <- factor(training_data$group)



set.seed(101422)


rf_model <- randomForest(x = predictors, 
                         y = response, 
                         ntree = 1000, 
                         importance = TRUE)


print(rf_model$confusion)


important_proteins_rf <- rf_model$importance %>%
  as_tibble() %>%
  mutate(protein = rownames(rf_model$importance)) %>%
  slice_max(MeanDecreaseGini, n = 20) %>%  # top 20
  pull(protein)


print(important_proteins_rf)

```

##### Logistic Regression

```{r}

library(caret)
library(dplyr)



train_data_for_model <- training_data %>%
  dplyr::select(-ados) %>%
  dplyr::mutate(group = as.factor(group))



train_control <- trainControl(method = "none")


logistic_model <- train(group ~ ., 
                        data = train_data_for_model,
                        method = "glm",
                        family = "binomial",
                        trControl = train_control)


logistic_coefficients <- coef(logistic_model$finalModel)


coeff_df <- as.data.frame(logistic_coefficients) %>%
  rownames_to_column(var = "protein") %>%
  filter(protein != "(Intercept)") %>%
  rename(coefficient = logistic_coefficients) %>%
  mutate(abs_coefficient = abs(coefficient))


top_20_proteins <- coeff_df %>%
  arrange(desc(abs_coefficient)) %>%
  slice(1:20) %>%
  pull(protein)


print(top_20_proteins)



```

##### Fuzzy Interaction

```{r}
all_top_proteins <- list(
  ttest = top_proteins_ttest,
  rf = important_proteins_rf,
  logreg = top_20_proteins
)


fuzzy_intersection_proteins <- all_top_proteins %>%
  unlist() %>%       
  table() %>% 
  .[. >= 2] %>%    
  names()            

print(fuzzy_intersection_proteins)

```

#### How are results affected by each modification?

```{r}
# Evaluate each set of selected proteins on testing data
evaluate_model_accuracy <- function(selected_proteins, model_name) {
  # Explicitly use dplyr::select() and tidyselect::all_of()
  predictors <- dplyr::select(testing_data, tidyselect::all_of(selected_proteins))
  response <- factor(testing_data$group)
  
  rf_test <- randomForest(x = predictors, y = response, ntree = 100)
  accuracy <- sum(diag(rf_test$confusion)) / sum(rf_test$confusion)
  
  cat("Accuracy of", model_name, "model on testing data:", accuracy, "\n")
}

# Evaluate individual methods
evaluate_model_accuracy(top_proteins_ttest, "T-test")
evaluate_model_accuracy(important_proteins_rf, "Random Forest")
#evaluate_model_accuracy(top_20_proteins, "Logistic Regression")

# Evaluate fuzzy intersection
evaluate_model_accuracy(fuzzy_intersection_proteins, "Fuzzy Intersection")

```

Each method yields different predictive performance, with Random Forest and the Fuzzy Intersection performing similarly well. The fuzzy intersection can be a useful compromise when seeking a balance between different selection criteria, but in this case, it does not significantly outperform Random Forest alone.

### Improved classifier

Accuracy of T-test model on testing data: 0.720524

Accuracy of Random Forest model on testing data: 0.7894737

Accuracy of Fuzzy Intersection model on testing data: 0.8241758

Each method shows varying levels of predictive accuracy, with Random Forest and the Fuzzy Intersection achieving comparable results. The fuzzy intersection provides a balanced approach by integrating selections from multiple criteria, though in this instance, it does not notably exceed the performance of Random Forest on its own.

```{r}
library(caret)


training_data$group <- factor(training_data$group)


control <- rfeControl(functions = lrFuncs, method = "cv", number = 10)


predictors_data <- training_data[, !names(training_data) %in% c("group", "ados")]


rfe_results <- rfe(
  x = predictors_data,
  y = training_data$group,
  sizes = c(5, 10, 15),
  rfeControl = control
)


simpler_panel <- predictors(rfe_results)
print(simpler_panel)

```

```{r}

evaluate_model_accuracy(simpler_panel, "Simpler Panel")

```

```{r}
alternative_proteins <- all_top_proteins %>%
  unlist() %>%       
  table() %>% 
  .[. >= 3] %>%    
  names()


evaluate_model_accuracy(alternative_proteins, "Alternative Panel")

```

-   **Initial Panels**: T-test, Random Forest, Logistic Regression, Fuzzy Intersection (original)

-   **New Panels**: Simpler Panel (RFE) and Alternative Panel (modified fuzzy intersection)
