

### 1.
Since `biomarker-raw.csv` contains protein concentration levels across various samples, let's examine the distribution of a sample of these protein values.
```{r}
library(ggplot2)
library(reshape2)
```

```{r}
data <- read.csv("data2/biomarker-raw.csv")

# Convert the selected protein columns to numeric (skip non-numeric values)
proteins <- data[, 3:7]  # Select a sample of 5 protein columns
proteins <- data.frame(lapply(proteins, function(x) as.numeric(as.character(x))))

# Melt the data for ggplot

proteins_long <- melt(proteins, variable.name = "Protein", value.name = "Level")

# Plot histograms for each selected protein
ggplot(proteins_long, aes(x = Level)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_wrap(~Protein, scales = "free") +
  labs(title = "Distribution of Raw Protein Levels (Sample Proteins)",
       x = "Protein Level", y = "Frequency") +
  theme_minimal()

```
The histograms reveal that the distributions of raw protein levels are skewed right, with some values extending to higher ranges. 
This skewness is a common reason to apply a log transformation, which can help normalize these distributions.

### 2.
```{r}
library(dplyr)
library(tidyr)
```

```{r}
# Load necessary libraries
library(dplyr)
library(tidyr)

# Load the data (assuming it's processed after removing outlier trimming)
data <- read.csv("data2/biomarker-raw.csv")

# Convert relevant columns to numeric, skipping metadata columns
protein_data <- data[, -c(1,2)]  # Adjust index based on the actual data structure

# Define a function to identify outliers based on ±3 standard deviations
identify_outliers <- function(x) {
  upper_limit <- mean(x, na.rm = TRUE) + 3 * sd(x, na.rm = TRUE)
  lower_limit <- mean(x, na.rm = TRUE) - 3 * sd(x, na.rm = TRUE)
  return(x < lower_limit | x > upper_limit)
}

# Apply the outlier function across all protein columns
outliers <- protein_data %>%
  mutate(across(everything(), identify_outliers)) %>%
  rowwise() %>%
  mutate(outlier_count = sum(c_across(everything()), na.rm = TRUE)) %>%
  ungroup()

# Merge with subject ID and group columns
outlier_summary <- data %>%
  select(Group) %>%
  bind_cols(outliers["outlier_count"])

# Summarize the number of outlying values per subject and by group
outlier_by_group <- outlier_summary %>%
  group_by(Group) %>%
  summarise(avg_outliers = mean(outlier_count, na.rm = TRUE),
            total_outliers = sum(outlier_count, na.rm = TRUE),
            subjects_with_outliers = sum(outlier_count > 0))

# View results
print(outlier_by_group)

```
#### Are there specific subjects (not values) that seem to be outliers?
Yes, there are specific subjects with outliers. According to the table of nsummary, there are 2 subjects with outlying values in the ASD group.

#### Are outliers more frequent in one group or the other?
Outliers are more frequent in the ASD group.
ASD group with a total of 2 outliers, while the TD group has none.

