
```{r}
library(ggplot2)
```


### 1.
Since `biomarker-raw.csv` contains protein concentration levels across various samples, let's examine the distribution of a sample of these protein values.
```{r}
library(ggplot2)
library(reshape2)
```

```{r}
data <- read.csv("data/biomarker-raw.csv")

# Convert the selected protein columns to numeric (skip non-numeric values)
proteins <- data[, 3:7]  # Select a sample of 5 protein columns
proteins <- data.frame(lapply(proteins, function(x) as.numeric(as.character(x))))

# Melt the data for ggplot

proteins_long <- melt(proteins, variable.name = "Protein", value.name = "Level")

# Plot histograms for each selected protein
ggplot(proteins_long, aes(x = Level)) +
  geom_histogram(bins = 30, color = "black", fill = "skyblue") +
  facet_wrap(~Protein, scales = "free") +
  labs(title = "Distribution of Raw Protein Levels (Sample Proteins)",
       x = "Protein Level", y = "Frequency") +
  theme_minimal()

```
The histograms reveal that the distributions of raw protein levels are skewed right, with some values extending to higher ranges. 
This skewness is a common reason to apply a log transformation, which can help normalize these distributions.



